# Chapter 3: Poisson Distribution {.unnumbered}

The **Poisson distribution** models the probability of a given number of events occurring in a fixed interval of time or space, assuming that the events occur independently and at a constant average rate.

The probability mass function is:

$$
P(X = k) = \frac{e^{-\lambda} \cdot \lambda^k}{k!}, \quad k = 0, 1, 2, ...
$$

Where:

-   $X$ is the number of events (e.g., target copies per droplet),\
-   $\lambda$ is the average number of events per interval (copy per droplet in ddPCR),\
-   $e$ is Euler’s number, approximately 2.71828.

**When to Use the Poisson Distribution**

Poisson distribution is appropriate under the following conditions:

-    Events are rare and occur independently.

-   The probability of two or more events happening simultaneously is negligible.

-   The average event rate ($\lambda$) is constant.

-   The number of possible trials is large, and the success probability is small.

**Poisson Approximation to Binomial**

When:

-   $n$ is large (e.g., \> 20),

-   $p$ is small (e.g., \< 0.01), and

-   $\lambda = n \cdot p$ is moderate (e.g., ≤ 5),

Then the binomial distribution $B(n, p)$ can be approximated by a poisson distribution with parameter $\lambda = np$.

**Example 1**: Rare Event Detection 

Probability that at least 4 sets of triplets are born, assuming a triplet birth probability of 0.0001 and 10,000 births:

**In R**:

```{r}
lambda <- 10000 * 0.0001  # λ = np = 1
1 - ppois(3, lambda)       # P(X ≥ 4)
```

**Example 2**: Defective Products

If 2% of products are defective, and we sample 100 products, what is the probability that exactly 3 are defective?

```{r}
lambda <- 100 * 0.02
dpois(3, lambda)
```

**Example 3**: Typos on a Page

If the number of typos per page follows Poisson(λ = 1), what's the probability of at least one typo?

```{r}
1 - dpois(0, 1) 
```

**Example 4**: No Accidents Today 

If the average number of accidents per day on a highway is λ = 3, the probability of no accidents today is:

```{r}
dpois(0, 3)
```

**Interpretation in ddPCR** 

In ddPCR, the number of target molecules per droplet follows a Poisson distribution with parameter $\lambda$. This is because each droplet is randomly and independently assigned zero or more DNA molecules. The key relationship used in ddPCR is:

$$
P(\text{no target  in  droplet}) = P(X=0) = e^{-\lambda}
$$ Which leads to the estimate:

$$
\lambda = -ln(\frac{\text{number of negative droplets}}{\text{total number of droplets}})
$$

This formula forms the foundation of **absolute quantification** in ddPCR, as we will explore in the next chapter.
