# Chapter 4: Estimating Copy Number in ddPCR {.unnumbered}

Digital PCR (ddPCR) partitions a reaction mix into thousands of droplets, each acting as an independent PCR microreaction.\
The goal is to estimate the **absolute concentration** of the target gene in the original sample, based on how many droplets are **positive** (i.e., contain at least one copy of the target sequence).

**Key Parameters**

-   $N$: total number of droplets\
-   $p$: number of positive droplets (detected signal)\
-   $v$: volume of a single droplet (in μL)\
-   $d$: dilution factor (from original sample to the PCR well)

**Estimating Copy Per Droplet (CPD)**

From Poisson theory, the probability that a droplet is negative (i.e., contains no target) is:

$$
P(X = 0) = e^{-\lambda}
$$

Rearranging:

$$
\lambda = -\ln\left(\frac{N - p}{N}\right) = -\ln\left(1 - \frac{p}{N}\right)
$$

This $\lambda$ is the **mean copy number per droplet**.

**Estimating Concentration (copies/μL)** 

We can convert copy per droplet (CPD) into **concentration** using the droplet volume $v$ and dilution factor $d$:

$$
C = \frac{\lambda \cdot d}{v} = -\frac{d}{v} \ln\left(1 - \frac{p}{N}\right)
$$

This gives the **absolute concentration (copies/μL)** in the original sample.

**Example 1**: Moderate Concentration 

-   Reaction volume: 20 μL\
-   Total droplets: 20,000\
-   Positive droplets: 200\
-   No dilution ($d = 1$)

```{r}
N <- 20000
p <- 200
v <- 20 / N
d <- 1
lambda <- -log(1 - p / N)
C <- lambda * d / v
paste(round(C, 3), "copies/uL")
```

**Example 2**: Rare Target Detection  

Only 1 positive droplet observed among 20,000:  

```{r}
p <- 1
lambda <- -log(1 - p / N)
C <- lambda / v
paste(round(C, 3), "copies/uL")
```
**Visualizing Positive Droplets vs. Concentration**  

```{r,echo=FALSE, fig.cap="Estimiated concentration vs positive droplet"}
library(ggplot2)

result <- data.frame(p.drop = 1:20000)
result$conc <- round(-1 / (20 / 20000) * log(1 - result$p.drop / 20000), 3)

ggplot(result, aes(x = p.drop, y = conc)) +
  geom_point(size = 0.1) +
  labs(title = "Estimated Concentration vs. Positive Droplets",
       x = "Positive Droplets (p)",
       y = "Estimated Concentration (copies/uL)")

```

**Summary**  

-   ddPCR uses the fraction of positive droplets to estimate $\lambda$  
-   $\lambda$ is then converted to copies per μL  
-   This method requires no standard curve, unlike qPCR  
-   Particularly useful for low copy number detection and absolute quantification

In the next chapter, we will explore how $\lambda$ relates to detection probability and how it informs **Limit of Detection (LoD)** and **Confidence Intervals** in ddPCR.

