# Chapter 5: CPD and Detection Probability {.unnumbered}

Once we estimate the **mean number of target copies per droplet** ($\lambda$), we can use the **Poisson distribution** to explore how the number of positive droplets is related to the actual concentration.

This chapter focuses on how **copy per droplet (CPD)** affects the probability of detecting the target DNA, particularly when the concentration is low.

**Poisson Probabilities for Different $k$ Values** 

Let’s consider $\lambda = 0.25$ (i.e., 0.25 copies per droplet), and a total of $n = 20,000$ droplets.

We compute the expected number of droplets with 0, 1, 2, ... copies using the Poisson formula:

$$
P(X = k) = \frac{e^{-\lambda} \lambda^k}{k!}
$$

```{r}
options(scipen = 1)
lambda <- 0.25
N <- 20000
k <- 0:5
expected <- dpois(k, lambda) * N
data.frame(k, expected)
```

Result:

| k   | Expected Droplets |
|-----|-------------------|
| 0   | 15,576            |
| 1   | 3,894             |
| 2   | 487               |
| 3   | 41                |
| 4   | 2.5               |
| 5   | 0.13              |

Most droplets are expected to contain 0 or 1 copy, confirming that low-λ conditions result in sparse detection.

**Probability of a Negative Droplet vs. CPD**

We can also explore how the probability of a negative droplet ($P(X = 0)$) decreases as CPD ($\lambda$) increases:

```{r}
cpd.list <- c(0.05, 0.1, 0.15, 0.2, 0.25,
              0.5, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

prob <- round(exp(-cpd.list), 7)
df_prob <- data.frame(CPD = cpd.list,
                      P_negative = prob)

df_prob
```

Example values:

| CPD (λ) | P(Negative Droplet) |
|---------|---------------------|
| 0.05    | 0.9512294           |
| 0.25    | 0.7788008           |
| 1       | 0.3678794           |
| 3       | 0.0497871           |
| 5       | 0.0067379           |
| 10      | 0.0000454           |

**Visualization** 

```{r, echo=FALSE,  fig.cap="Probability of negative droplet vs cpd"}
library(ggplot2)

ggplot(df_prob, aes(x = CPD, y = P_negative)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Probability of Negative Droplet vs. CPD",
       x = "Copy Per Droplet (λ)",
       y = "P(Negative Droplet)")
```

We also can define the **Limit of Detection (LoD)** as the lowest CPD (λ) where the probability of a negative droplet falls below a threshold (e.g., 5%).

```{r}
alpha <- 0.05  # maximum acceptable false negative probability
lambda_seq <- seq(0, 5, by = 0.01)
p0 <- exp(-lambda_seq)
lod_lambda <- lambda_seq[which(p0 <= alpha)[1]]
lod_lambda
```

This means that at least 3 copies per droplet are needed to ensure that the probability of getting no signal at all is below 5%.

```{r, echo =FALSE, fig.cap="lod threshold"}
plot(lambda_seq, p0, type = "l", col = "blue",
     ylab = "P(X = 0)", xlab = "Lambda (CPD)",
     main = "LoD Threshold for P(X = 0) ≤ 0.05")
abline(h = alpha, col = "red", lty = 2)
abline(v = lod_lambda, col = "darkgreen", lty = 2)
text(lod_lambda, alpha + 0.05, paste0("LoD ≈ ", lod_lambda), pos = 4)
```

**Interpretation**

-   When $\lambda$ is small (e.g., 0.05), almost all droplets are negative.\
-   As $\lambda$ increases, the probability of a droplet containing at least one copy increases rapidly.\
-   At $\lambda = 3$, only \~5% of droplets are negative.

This relationship is critical when defining detection thresholds.

**Application to ddPCR LoD**

One way to define Limit of Detection (LoD) is to set a minimum $\lambda$ such that:

$$  
P(\text{No Detection}) = P(X=0) ≤ \alpha
$$ For example:

If we define $\alpha = 0.05$, then LoD occurs around $\lambda = 3$.

**Summary**  

-   The Poisson model connects CPD with the likelihood of detecting target molecules.

-   This is essential for **designing sensitive assays**, particularly when working with **low-abundance targets**.

-   Next, we will explore how to calculate **confidence intervals** for $\lambda$ and the resulting copy number.
