# Chapter 6: Confidence Intervals and Limit of Detection (LoD) {.unnumbered}

In ddPCR, it is important not only to estimate the concentration of the target gene, but also to quantify the **uncertainty** of that estimate. This is typically done by calculating **confidence intervals (CIs)** for the estimated copy number.

**Confidence Interval for Proportion of Positive Droplets**

Let:

-   $N$: total number of droplets

-   $p$: number of positive droplets

Then the observed proportion is:

$$
\hat{p} = \frac{p}{N}
$$

The confidence interval for this proportion can be calculated using binomial methods, such as:

-   **Wald method** (simplest, but less accurate when $p$ is near 0 or 1)\
-   **Wilson score interval**\
-   **Exact (Clopper-Pearson) method**

We can then transform this interval into a CI for $\lambda$ (CPD), and then into concentration.

**R Example**: Using `binom` Package

```{r}
library(binom)

N <- 20000
p <- 200
p_hat <- p / N

binom.confint(p, N, methods = "exact")
```

This gives the lower and upper bounds for the proportion of positive droplets.

We then compute:

```{r}
lambda_lower <- -log(1 - 0.008589)
lambda_upper <- -log(1 - 0.010316)

v <- 20 / N  # droplet volume
d <- 1       # no dilution
conc_lower <- lambda_lower * d / v
conc_upper <- lambda_upper * d / v

round(c(conc_lower, conc_upper), 2)
```

**Interpretation**

-   The estimated concentration is about **10.05 copies/μL**

-   The 95% confidence interval is roughly **[8.6, 10.8] copies/μL**

-   The uncertainty reflects both the number of positive droplets and the total droplet count

**Defining Limit of Detection (LoD)**

**LoD** is often defined as the lowest concentration that gives a statistically significant number of positive droplets, with an acceptably low false negative rate.

A common approach is:

-   Simulate or calculate the probability of detecting at least one positive droplet at a given $\lambda$
-   Require that this probability exceeds 1 - α, e.g., 95%

This was demonstrated in the previous chapter using:

$$
P(X=0) = e^{-\lambda}, \ \lambda_{LOD} = -ln(\alpha)
$$

**Summary** 

-   Confidence intervals for ddPCR results can be constructed by first computing the CI for the proportion of positive droplets.
-   The LoD can be estimated as the smallest concentration that still results in a sufficiently high probability of detection.

This closes the foundational mathematical modeling of ddPCR.
