# Chapter 7: Applications of ddPCR in Genomic Analysis {.unnumbered}

Digital PCR (ddPCR) enables absolute quantification of nucleic acids without the need for calibration curves.\
This makes it a powerful tool in genomic applications where precision is critical, such as:

-   **Copy Number Variation (CNV) detection**
-   **Rare allele or mutation detection**

**Copy Number Variation (CNV) Detection** 

**Copy Number Variation (CNV)** refers to a gain or loss of gene copies in the genome.
ddPCR quantifies the number of gene copies in a sample and compares it to a **reference gene** with known normal copy number (usually 2).

The relative copy number is calculated as:

$$
\text{Copy Ratio} = \frac{\text{Target Gene Concentration}}{\text{Reference Gene Concentration}}
$$

**Example**:

```{r}
target <- 10.1 # estimated concentration of target gene (copies/μL)
reference <- 5.0 # estimated concentration of reference gene (copies/μL)
copy_ratio <- target / reference
round(copy_ratio, 2)
```

**Interpretation**

| Copy Ratio | Interpretation        |
|------------|-----------------------|
| \~1.0      | Normal diploid copy   |
| \~0.5      | Heterozygous deletion |
| \~2.0      | Gene duplication      |

ddPCR is especially useful for detecting subtle CNVs that may not be distinguishable by qPCR or arrays.  


**Rare Allele Detection** 

ddPCR can detect low-frequency mutations, even at < 0.1% mutant allele frequency (MAF). This is crucial for applications such as:

-   Detecting resistance mutations in cancer (e.g., EGFR T790M)
-   Tracking minimal residual disease (MRD)
-   Analyzing circulating tumor DNA (ctDNA)

**Dual-probe strategy**

Each droplet is tested for both:

-   Wild-type (WT) using one probe (e.g., HEX)
-   Mutant (MT) using another probe (e.g., FAM)
-   The Mutant Allele Frequency (MAF) is calculated as:

$$
\text{MAF} = \frac{\text{Mutanat Copies}}{\text{Mutant Copies + Wild-Type Copies}}
$$

**Example**:

```{r}
mutant <- 6
wt <- 994

maf <- mutant / (mutant + wt)
round(maf * 100, 3)
```

This demonstrates ddPCR’s ability to detect rare mutations in a background of wild-type DNA.

**Summary**

| Application           | Measurement                     | ddPCR Advantage                                  |
| --------------------- | ------------------------------- | ------------------------------------------------ |
| CNV Detection         | Copy ratio (Target / Reference) | Absolute quantification without a standard curve |
| Rare Allele Detection | Mutant Allele Frequency (MAF)   | Sensitive detection of <0.1% MAF                 |



